Differential effects of knapping skill on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-01-30 manuscript/manuscript.Rmd
incooperating Dietz's revision advice.

---
title: "Differential cultural reproduction of knapping skill and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment. We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper incorporates these two perspectives into a broader conceptual framework of cultural reproduction and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

@@

Recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations. From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

>>>>

---
title: "Differential effects of knapping skill on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment. We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].

Here we used experimental data from a multidisciplinary study of handaxe-making skill acquisition [@pargeter2019; @pargeter2020; @bayani2021] to investigate the interaction between learning processes and the reproduction of specific morphological targets (c.f. "mental template"). This investigation was motivated by the theoretical expectation that, just as developmental processes can act to channel biological variation and shape evolutionary trajectories [@laland2015], learning challenges might influence cultural evolutionary processes [e.g., @schillinger2014]. Results allowed us to identify particular aspects of handaxe morphology that are more and less constrained by learning difficulty, thus helping to partition sources of morphological variation [@lycett2015]. By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach assess the presence and nature of culturally reproduced mental templates. This complements previous work investigated reduction intensity [@shipton2015] and raw material form [@garcía-medrano2019] as alternative explanations for morphological variation and standardization at Boxgrove.

@@

Current conceptions of a "mental template" are thus more nuanced than the idea of a fully specified image in the mind of the maker that is directly expressed in material form and transmitted between minds. For example, Hutchence and Scott [-@hutchence2021], leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations. From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

In short, contemporary approaches to the concept of a mental template emphasize the causal importance of production process and constraints and the interaction between individual and group level phenomena. We again note the striking similarity of the perspectives to the concept of "constructive development" as a source of guided variation in evolution biology [@laland2015]. We thus sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.
